---
title: "Motif Enrichment in SEMplR (meSEMplR)"
author: "Jez Burgess"
date: "2025-07-09"
output:
  revealjs::revealjs_presentation:
    slide_level: 2
    incremental: true
---

## 1. Overview

- **Goal**: For a given list of differential ids, map to GENCODE data, define promoter regions, and establish foreground and backgroiund sets for motif enrichment analysis.
- **Key steps**:
  1. Parse & tidy GENCODE GTF - parseGENCODEgtf()
  2. Annotate with metadata - annotateGTF()
  3. Map user gene list to identify id type - computeMatchStats()
  4. Define enrichment pools - mapUserList()
  5. Establish background elements (matched/random) - defineBckgroundElements()
  [6. Extract sequences for foreground and background elements - getPromoterSeqs()]

---

## 2. Step 1: parseGENCODEgtf()

**Input**:
- GENCODE GTF file path (e.g. `gencode.v48.annotation.gtf.gz`)
- `feature_types = c("gene", "transcript")`

**Function call**:

```r
gtf <- parseGENCODEgtf(
  gtf_path,
  feature_types = c("gene", "transcript")
)
```

```{r show-gtf-head, results='hold'}
library(knitr)
# Example output (uncomment after defining gtf)
# kable(head(gtf)[, c("seqnames", "start", "end", "feature_type")],
#       caption = "First 6 rows of parsed GTF")
```

---

## 3. Step 2: annotateGTF()

**Input**:
- `gtf` (from previous slide)
- External metadata URLs or local files

**Function call**:

```r
gtf_map <- annotateGTF(
  gtf,
  metadata_urls = list(
    entrez = "path/to/Entrez.gz",
    hgnc   = "path/to/HGNC.gz",
    refseq = "path/to/RefSeq.gz"
  )
)
```

**Output**:
- `gtf_map` includes `flag_*` columns (e.g. `flag_Ensembl_canonical`)
- Stripped ID versions (`gene_id_base`, `transcript_id_base`)
- Joined `entrez_id`, `HGNC_symbol`, etc.

---

## 4. Step 3: compute_match_stats()

**Purpose**: Identify which column best matches your `user_list`.

**Function call**:

```r
stats <- compute_match_stats(
  user_list,
  map_df = gtf_map
)
```

```{r show-stats-head, results='hold'}
# Example output (uncomment after computing stats)
# kable(head(stats), caption = "Mapping statistics by ID column")
```
